Measuring neutron star tidal deformability with Advanced LIGO: a Bayesian analysis of mixed binary observations

### 2016-05-24 Paper/paper.tex
polished abstrsact a bit

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}

@@

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the star's tidal distortion by
the companion black hole (BH) during inspiral, and (b) its tidal disruption
(if that happens) near merger.
%
In this paper, we present the first fully Bayesian study of the measurability 
of neutron star (NS) tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$ from NSBH
observation(s), where we
take into account the crucial effect of BH spins as well.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\simeq30$.
% 
Second, we find that if we are to get a loud signal with SNR $\gtrsim 23$, 
we can put a factor of two bound on $\lambdans$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) in the same
number of observations, the associated $90\%$ confidence intervals shrink to
$\pm 50\%$ of the true value.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories completed their first observing run ``O1''
early-2016, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, they made the first terrestrial observation of gravitational 
waves~\cite{LIGOVirgo2016a}. Emitted by a pair of coalescing black holes, these
waves heralded an era of observational GW astrophysics as they traveled 
through Earth.
% 
By the end of the decade, we expect aLIGO to reach its design sensitivity,
as well as for Advanced Virgo and KAGRA to begin observing. 
% 
By $2018-19$, we expect the aLIGO instruments to reach their design
sensitivity, at which point we may expect to observe $\sim 70$ binary mergers a
year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf}. An important sub-class of these sources are NSBH
binaries, of which we expect $\mathcal{O}(10)$ GW observations every 
year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.

>>>>

\usepackage[colorlinks, pdfborder={0 0 0}, plainpages=false]{hyperref}
\usepackage{graphicx}
\usepackage{xspace}
\usepackage[usenames,dvipsnames]{color}
\usepackage{amssymb}
\usepackage[dvipsnames]{xcolor}
\usepackage{placeins}
\usepackage{lineno}
\linenumbers

@@

\begin{abstract}
The pioneering discovery of gravitational waves (GW) by Advanced LIGO has ushered
us into an era of observational GW astrophysics. Compact binaries remain the 
primary target sources for GW observation, of which neutron star - black hole 
(NSBH) binaries form an important subset.
% 
GWs from NSBH sources carry signatures of (a) the star's tidal distortion by
the companion black hole (BH) during inspiral, and (b) its tidal disruption
(if that happens) near merger.
%
In this paper, we present a Bayesian study of the measurability of
neutron star (NS) tidal deformability $\lambdans\propto (R/M)_\mathrm{NS}^{5}$
from GW observation(s) of disruptive NSBH coalescences, taking into account the
crucial effect of BH spins.
% 
First, we find that if BHBH templates are used to estimate source 
parameters for an NSBH signal, a significant bias is introduced in the 
inference if the signal-to-noise ratio (SNR) is greater than $\simeq30$.
% 
Second, we find that if we are to get a loud signal with SNR $\gtrsim 23$, 
we can put a factor of two bound on $\lambdans$.
% 
Finally, we study the improvement in our measurement of $\lambdans$ from 
multiple observations. Using an approximately astrophysical population, 
we find that (i) the median measured $\lambdans$ catches up to within
$10\%$ of the true value in approximately $20$ observations; (ii) in the same
number of observations, the associated $90\%$ confidence intervals shrink to
$\pm 50\%$ of the true value.
% 
We find these results encouraging, and recommend that an effort to 
measure $\lambdans$ be taken within the LIGO-Virgo Collaboration once
we begin to detect NSBH coalescences.
\end{abstract}

@@

The Advanced LIGO (aLIGO) observatories completed their first observing run ``O1''
early-2016, operating at a factor of $3-4$ higher gravitational-wave
(GW) strain sensitivity than their first-generation 
counterparts~\cite{Shoemaker2009}.
% 
During O1, they made the first terrestrial observation of gravitational 
waves~\cite{LIGOVirgo2016a}. Emitted by a pair of coalescing black holes, these
waves heralded an era of observational GW astrophysics as they traveled 
through Earth.
% 
By the end of the decade, we expect aLIGO to reach its design sensitivity,
as well as for Advanced Virgo and KAGRA to begin observing. 
% 
At its design sensitivity, we expect aLIGO to observe $\sim 70$ compact binary
mergers a year~\cite{Abadie:2010cf}.

Coalescing binaries of compact objects, i.e. black holes (BH) and/or
neutron stars (NS) are the primary target sources for 
aLIGO~\cite{Abbott:2016nhf,LIGOVirgo2016a}. An important sub-class of these
are NSBH binaries, of which we expect $\mathcal{O}(10)$ GW observations
every year~\cite{Abadie:2010cf}.
NSBH binaries are of interest for multiple reasons. For instance,
they have been long associated with (as possible progenitors of) short
Gamma-ray Bursts (GRBs)~\cite{eichler:89,1992ApJ...395L..83N,moch:93,Barthelmy:2005bx,
2005Natur.437..845F,2005Natur.437..851G,Shibata:2005mz,Paschalidis2014,
Tanvir:2013}. Under favorable conditions, the NS can get disrupted by
the tidal field of its companion BH. Once disrupted, most of the NS
material falls into the BH over an $\mathcal{O}(1$ms$)$ time-scale,
with the rest forming an accretion disk around the BH. This disk-BH system,
which survives for $0.1-1$s, is hypothesized to drive SGRBs through the
production of relativistic jets~\cite{Foucart:2015a,
Lovelace:2013vma,Deaton2013,Foucart2012,Shibata:2005mz,Paschalidis2014}.
A {\it coincident} detection of both gravitational and electromagnetic
waves (gamma rays) from an NSBH system is therefore a {\it unique}
possibility to confirm this hypothesis~\cite{Abbott:2016wya}.
% 
% Most of the matter falls into the BH in ~1ms. But the disk itself has an
% accretion timescale of ~100ms-1s. The disk-BH system is the expected engine
% for a SGRB, through the production of a relativistic jet. The unbound material,
% on the other hand, is responsible for an EM signal in the infrared/optical, 
% powered by radioactive decays in the ejecta. And possibly a radio afterglow
% through interactions with the ISM. But not gamma-rays (unless you are referring
% to the very small amount of material ejected in the jet, but that's not what
% people usually mean when talking about material unbound by the merger).
% 
Another question that NSBH mergers can help answer is `what is the 
nature of matter at nuclear densities', i.e. what is the equation of state
(EoS) of neutron star material. This is what we will focus on in this paper.
% 
GWs emitted by NSBH coalescences carry subtle hints of the material structure
of NSs. During early inspiral, the (weak) tidal field of the companion BH
deforms the star, thereby exciting its resonant oscillation modes. The induced
quadrupolar moment $Q_{ij}$ of the star depends on the
tidal field $\mathcal{E}_{ij}$ of its companion through the relation
$Q_{ij} = -\lambda\mathcal{E}_{ij}$, where $\lambda$ is an EoS-dependent
tidal deformability parameter that is related to NS's dimensionless Love number
$k_2$ and radius $R_\mathrm{NS}$ as $\lambda:=\frac{2}{3G}k_2 R_\mathrm{NS}^5$.
The coherence of GW emission between stellar quadrupolar waves and orbital
waves drains energy from the orbit and increases the inspiral rate~\cite{Flanagan2008}.
We know the leading and next-to
-leading order terms in post-Newtonian (PN) theory that capture this 
effect~\cite{Vines2011}, entering binary phasing at $5$PN order. As the NS gets
disrupted close to binary's merger, the quadrupole moment of the system falls
monotonically. As a result, the emitted GWs are subdued significantly
post-disruption (as compared to what they would be for a 
BBH)~\cite{Kyutoku:2010zd,Lackey:2013axa,Lovelace:2013vma,Foucart:2015a,
Pannarale:2015jia}.
